---
title: 'Lab 10: Tidyverse I: Pipes and Dplyr'
output: pdf_document
---

```{r, include=FALSE}
knitr::opts_chunk$set(cache=TRUE, autodep=TRUE, cache.comments=TRUE)
```

Name: Rufus Petrie    

**This week's agenda**: learning to master pipes and `dplyr`.

```{r}
# Load the tidyverse!
library(tidyverse)
```

Pipes to base R
===

For each of the following code blocks, which are written with pipes, write equivalent code in base R (to do the same thing).

- **1a.** 

```{r}
# Pipes:
letters <- "abcde"
letters %>%
  toupper %>%
  paste(collapse="+") 

# Base R:
paste(toupper(letters), collapse="+")
```

- **1b.** 

```{r}
# Pipes:
phrase <- "     Ceci n'est pas une pipe     "
"     Ceci n'est pas une pipe     " %>% 
  gsub("une", "un", .) %>%
  trimws

# Base R:
trimws(gsub("une", "un", phrase))
```

- **1c.**

```{r}
# Pipes:
rnorm(1000) %>% 
  hist(breaks=30, main="N(0,1) draws", col="pink", prob=TRUE) 

# Base R:
hist(rnorm(1000), breaks=30, main="N(0,1) draws", col="pink", prob=TRUE)
```

- **1d.** 

```{r}
# Pipes:
rnorm(1000) %>% 
  hist(breaks=30, plot=FALSE) %>%
  "[["("density") %>%
  max

# Base R:
max(hist(rnorm(1000), breaks=30, plot=FALSE)$density)
```

Base R to pipes
===

For each of the following code blocks, which are written in base R, write equivalent code with pipes (to do the same thing).

- **2a.** Hint: you'll have to use the dot `.`, as seen above in Q1b, or in the lecture notes.

```{r}
# Base R:
paste("Your grade is", sample(c("A","B","C","D","R"), size=1))

# Pipes:
c("A","B","C","D","R") %>% 
  sample(size=1) %>% 
  paste("Your grade is", .)
```

- **2b.** Hint: you can use the dot `.` again, in order to index `state.name` directly in the last pipe command.

```{r}
# Base R: 
state.name[which.max(state.x77[,"Illiteracy"])] 

# Pipes:
state.x77 %>% 
  .[,"Illiteracy"] %>% 
  which.max %>% 
  state.name[.]
```

- **2c.** Hint: if `x` is a list of length 1, then `x[[1]]` is the same as `unlist(x)`.

```{r}
str.url = "http://www.stat.cmu.edu/~ryantibs/statcomp/data/trump.txt"

# Base R:
lines = readLines(str.url)
text = paste(lines, collapse=" ")
words = strsplit(text, split="[[:space:]]|[[:punct:]]")[[1]]
wordtab = table(words)
wordtab = sort(wordtab, decreasing=TRUE)
head(wordtab, 10)

# Pipes:
readLines(str.url) %>% 
  paste(collapse=" ") %>% 
  strsplit(split="[[:space:]]|[[:punct:]]") %>% 
  unlist %>% 
  table %>%
  sort(decreasing = TRUE) %>% 
  head(10)
```

- **2d.** Hint: the only difference between this and the last part is the line `words = words[words != ""]`. This is a bit tricky line to do with pipes: use the dot `.`, once more, and manipulate it as if were a variable name.

```{r}
# Base R:
lines = readLines(str.url)
text = paste(lines, collapse=" ")
words = strsplit(text, split="[[:space:]]|[[:punct:]]")[[1]]
words = words[words != ""]
wordtab = table(words)
wordtab = sort(wordtab, decreasing=TRUE)
head(wordtab, 10)

# Pipes:
readLines(str.url) %>% 
  paste(collapse=" ") %>% 
  strsplit(split="[[:space:]]|[[:punct:]]") %>% 
  unlist %>% 
  .[.!=""] %>% 
  table %>% 
  sort(decreasing = TRUE) %>% 
  head(10)
```

Sprints data, revisited
===

Below we read in a data frame `sprint.w.df` containing the top women's times in the 100m sprint, as seen in previous labs. We also define a function `factor.to.numeric()` that was used in Lab 8, to convert the Wind column to numeric values. In what follows, use `dplyr` and pipes to answer the following questions on `sprint.w.df`.

```{r}
sprint.w.df = read.table(
  file="http://www.stat.cmu.edu/~ryantibs/statcomp/data/sprint.w.dat",
  sep="\t", header=TRUE, quote="", stringsAsFactors=TRUE)

factor.to.numeric = Vectorize(function(x) {
  x = strsplit(as.character(x), split = ",")[[1]]
  ifelse(length(x) > 1, 
         as.numeric(paste(x, collapse=".")), 
         as.numeric(x))
})
```

- **3a.** Convert the Wind column to numeric using `factor.to.numeric()`. Hint: use `mutate_at()`, and reassign `sprint.w.df` to be the output.

```{r}
sprint.w.df <- sprint.w.df %>% 
  mutate_at("Wind", factor.to.numeric)
```

- **3b.** Run a linear regression of the Time on Wind columns, but only using data where Wind values that are nonpositive, and report the coefficients. Hint: use `filter()`, and use the dot `.` to pipe into the `lm()` function appropriately.

```{r}
sprint.w.df %>% 
  filter(Wind<0) %>% 
  lm(Time ~ Wind, data = .) %>% 
  coef
```

- **3c.** Plot the Time versus Wind columns, but only using data where Wind values that are nonpositive, and label the axes appropriately. Hint: recall that for a data frame, with columns `colX` and `colY`, you can use `plot(colY ~ colX, data=df)`, to plot `df$colY` versus `df$colX`.

```{r}
sprint.w.df %>% 
  filter(Wind<0) %>% 
  plot(Time ~ Wind, data=., main="Time vs. Windspeed", xlab="Windspeed", ylab="Time")
```

- **3d.** Reorder the rows in terms of increasing Wind, and then display only the women who ran at most 10.7 seconds. Hint: do this with one single flow of pipe commands; use `arrange()`, `filter()`.

```{r}
sprint.w.df %>% 
  arrange(Wind) %>% 
  filter(Time<=10.7)
```

- **3e.** Now reorder the rows in terms of increasing Time, and *then* increasing Wind, and again display only the women who ran at most 10.7 seconds, but only display the Time, Wind, Name, and Date columns. Hint: a single flow of pipe commands will do; note that `arrange()` can take multiple columns that you want to sort by, and the order you pass them specifies the priority.

```{r}
sprint.w.df %>% 
  arrange(Time, Wind) %>% 
  filter(Time <= 10.7) %>% 
  select(Time, Wind, Name, Date)
```

Prostate cancer data, revisited
===

Below we read in a data frame `pros.df` containing measurements on men with prostate cancer, as seen in previous labs. As before, in what follows, use `dplyr` and pipes to answer the following questions on `pros.df`.

```{r}
pros.df = 
  read.table("http://www.stat.cmu.edu/~ryantibs/statcomp/data/pros.dat")
```

- **4a.** Among the men whose lcp value is equal to the minimum value, report the lowest and highest lpsa score.

```{r}
pros.df %>% 
  filter(lcp==min(lcp)) %>% 
  filter(lpsa==min(lpsa) | lpsa==max(lpsa))
```

- **4b.** Order the rows by decreasing age, then decreasing lpsa score, and display the rows from men who are older than 70, but only the age, lpsa, lcavol, and lweight columns.

```{r}
pros.df %>% 
  arrange(desc(age), desc(lpsa)) %>% 
  filter(age>70) %>% 
  select(age, lpsa, lcavol, lweight)
```

- **4c.** Run a linear regression of the lpsa on lcavol and lweight columns, but only using men whose lcp value is strictly larger than the minimum value, and report a summary of the fitted model.

```{r}
pros.df %>% 
  filter(lcp>min(lcp)) %>% 
  lm(lpsa ~ lcavol + lweight, data = .) %>% 
  summary
```

- **4d.** Extend your code in the last part, still just using a single flow of pipe commands in total, to extract the p-values associated with each of the coefficients in the fitted model.

```{r}
pros.df %>% 
  filter(lcp>min(lcp)) %>% 
  lm(lpsa ~ lcavol + lweight, data = .) %>% 
  summary %>% 
  .$coefficients %>% 
  .[,"Pr(>|t|)"]
```